\begin{abstract}
Modern compute-intensive software is written against a rapidly changing ecosystem of accelerators, programming APIs, compiler stacks, and portability layers. As hardware and software platforms evolve, kernels in AI, scientific computing, simulation, graphics, and data-intensive workloads often need to migrate across CUDA, OpenMP, OpenCL, OpenMP target offload, and related parallel APIs. Large language models and autonomous coding agents are increasingly proposed as tools for this migration, but the field still lacks a reliable way to measure whether they can preserve the low-level parallel semantics that make such translations correct, including thread indexing, synchronization, memory management, host--device coordination, and API-specific execution structure.

We present \parbench{}, a kernel-centric benchmark framework designed to isolate and measure LLM-based parallel API translation under executable, reproducible conditions. Unlike general code benchmarks that are largely sequential, parallel-code benchmarks that emphasize generation from natural-language prompts, or repository-level studies that confound translation with build-system reconstruction, \parbench{} fixes the surrounding build, run, and verification infrastructure through declarative benchmark specifications and asks models to translate only the computational kernels. The benchmark draws on multiple open-source HPC suites to cover representative cross-API translation directions among CUDA, OpenMP, OpenCL, and OpenMP target offload. To probe whether success reflects robust translation rather than surface-form memorization, \parbench{} also includes AST-driven semantics-preserving source augmentation. We further evaluate \parbench{} on state-of-the-art open and proprietary LLMs, showing that it exposes persistent barriers to reliable parallel code translation, including direction asymmetry, multi-file coordination, incomplete API adaptation, and uneven robustness to source-level perturbations.
\end{abstract}

% \begin{abstract}
% We present \parbench{}, a kernel-centric benchmark framework for evaluating LLM-based parallel code translation. \parbench{} combines declarative benchmark specifications, an end-to-end build-run-verify harness, survey-grounded corpus construction from 35~open-source HPC repositories, and AST-driven semantics-preserving source augmentation. The corpus presented in this paper contains 96~specifications (87 verified-pass, 9 excluded for platform incompatibilities) over 35~kernels from five benchmark suites, yielding six standard translation directions plus four OpenMP Offload case-study directions.

% % src: paper_data_*.json > passk_campaign > aggregate_passk; statistical_analysis.json > model_comparison > pairwise[0]
% Under stochastic sampling (temperature~0.7, three independent attempts per task), pass@1 ranges from 23.9\% (\qwenshort{}) to 62.7\% (shared by \gptnew{} and \codex{}) over 142~unique source-target pairs, with the code-specialized \codex{} performing indistinguishably from the general-purpose \gptnew{} ($p = 1.0$, OR\,=\,0.93). Build-stage adaptation failures account for 50.0\% of \qwenshort{} L0 outcomes, making incomplete API-surface mapping the dominant bottleneck. On L0-conditional augmentation subsets, pass rates show no evidence of memorization-dependent degradation for the GPT models. \parbench{} exposes both current capability and the structural barriers---direction asymmetry, multi-file coordination, and incomplete API adaptation---that remain for reliable parallel code translation.
% \end{abstract}