\section{Discussion, Limitations, and Conclusion}
\label{sec:discussion}

% src: cross_model_comparison.json, quantitative_findings_qwen.json, quantitative_findings_gpt54.json
% src: statistical_analysis.json > omnibus_chi2; cross_model_comparison_*.json
\parbench{} shows that, when repository reconstruction is fixed, current LLMs can produce declared-oracle-passing translations for a non-trivial fraction of kernel-level parallel API tasks, but success remains highly structured by API direction, failure stage, and model tier. The primary bottleneck is build-stage API adaptation, especially when translations must introduce explicit device-memory and runtime structure. A task-level omnibus test on 142 L0 tasks ($\chi^2(2) = 44.4$, $p < 10^{-10}$, Cram\'er's $V = 0.32$) detects significant heterogeneity across three models, where each task is classified as pass-any (at least one of three samples passes) or all-fail; pairwise analysis reveals two tiers: \gptnew{} and \codex{} achieve near-identical task-level pass rates (99/142 and 97/142; Fisher's $p = 1.0$), while both surpass \qwenshort{} by $4.2\times$ odds (Cohen's $h = 0.71$). Because the providers use unmatched sampling conditions (Section~\ref{sec:sampling-config}), the Qwen--GPT gap cannot be causally attributed to model capability alone; the within-provider comparison controls for this confound. Full pairwise statistical details appear in Appendix~\ref{sec:appendix-e4}.

The tier gap is one of determinism, not marginal accuracy: GPT-family models either solve a task reliably across all three samples or fail every attempt, whereas Qwen's failures are equally systematic but far more frequent (Section~\ref{sec:passk-analysis}). All three models share the same dominant bottleneck---incomplete API-surface adaptation at build time---but the GPT family resolves the majority of these, leaving residual build failures concentrated in the hardest directions (Section~\ref{sec:failure-taxonomy}). Direction rankings are consistent across models: removing explicit parallel constructs (CUDA-to-OpenMP) is substantially easier than introducing them (OpenCL-to-CUDA), suggesting that the difficulty gradient tracks API-surface complexity rather than model-specific memorization (Section~\ref{sec:direction-analysis}). Augmentation robustness reinforces this interpretation: GPT-family pass rates are stable across L1--L4, while \qwenshort{}'s declining augmentation rates suggest greater reliance on surface-form patterns, but do not reach significance at $\alpha = 0.05$ (Section~\ref{sec:augmentation-robustness}).

\textbf{Limitations.} The benchmark is correctness-focused: it does not measure speedups, occupancy, memory bandwidth, or efficiency. A harness-passing translation may still be unusably slow, a concern also raised by TRACE~\cite{TRACE2026}. The corpus is finite; some directions have small paired samples. The L0-conditional augmentation subset ranges from 3--22 kernels per model per standard direction (3--8 for omp\_target directions), with the smallest cells reflecting \qwenshort{}'s low L0 pass rate, but the balanced cross-model comparison is limited to 12 common kernels in a single direction. Small per-level sample sizes and the L0-conditioned selection of augmentation subsets complicate formal trend interpretation. Ten specs across six kernels were downgraded from strong or medium oracles to weak (stdout pattern plus exit code): six due to cross-API floating-point reduction-order divergence (cfd, hotspot, myocyte) and four due to synthesis asymmetry between reference implementations (bfs, nw, nn). This limits oracle strength but correctly avoids penalizing faithful translations. All results are from a single platform (NVIDIA RTX~4070, Ubuntu~24.04, HPC SDK~24.3); for GPU and offload code, compiler and runtime behavior varies across NVIDIA architecture generations and toolchain versions, so results are reproducible on this pinned platform but cross-platform generality remains future work. The stochastic evaluation (temperature~0.7, three samples) captures variability but does not measure iterative repair or agentic translation strategies. The three models use different reasoning mechanisms and sampling configurations (Section~\ref{sec:sampling-config}). The Qwen--GPT gap is large (task-level Cohen's $h = 0.71$) but cannot be fully attributed to model capability given unmatched sampling conditions. The within-provider \gptnew{}--\codex{} comparison controls for provider-side differences but both share the same Azure OpenAI infrastructure, limiting generalizability beyond this provider. \codex{}'s code specialization targets general software engineering tasks; the null result for parallel translation may not generalize to models specifically fine-tuned on HPC code. Appendix~\ref{sec:appendix-k} provides a structured Evaluation Card and reporting protocol; the artifact (Appendix~\ref{sec:artifact-availability}) documents per-record outputs that support reproduction and extension.

% \textbf{Future work.} Benchmark longevity depends on headroom. Both GPT models achieve pass@1\,=\,62.7\% at L0, so the current 142-task corpus is not yet saturated, but rapid model improvement will narrow that margin. Three extension mechanisms are available: adding kernels through the declarative spec format (JSON plus build recipe, no framework code), composing augmentation transformations beyond L4, and upgrading weak oracles to numeric verification to catch currently undetected semantic errors.

% Additional directions include self-repair experiments using the framework's built-in feedback loop, evaluation of agentic systems such as LASSI~\cite{LASSI2024} and ParaCodex~\cite{ParaCodex2026}, fine-tuned HPC models~\cite{HPCCoderV2}, efficiency-aware evaluation following TRACE~\cite{TRACE2026}, and expansion of the augmentation campaign to more kernels and directions.

% \textbf{Artifact Availability.}\parbench{} will be released under the MIT license on GitHub, with a versioned snapshot archived on Zenodo to provide a persistent DOI for long-term citability. The release includes all 206 kernel specifications, the build-run-verify harness, the evaluation and augmentation pipelines, and per-record result JSONs for all three models. Each kernel specification embeds its provenance: the upstream suite, repository URL, and pinned commit hash (Section~\ref{sec:suite-selection}). The five upstream suites are redistributed under their original licenses: Rodinia (BSD-like), XSBench (MIT), RSBench (MIT), HeCBench (MIT), and mixbench (GPL-3.0); all permit academic use and redistribution. A datasheet following \citet{Datasheets2021} accompanies the release. The declarative JSON spec format (Appendix~\ref{sec:appendix-a-schema}) serves as structured per-kernel metadata; Croissant export for ML dataset registries is planned as a post-publication extension.

% Appendix~\ref{sec:appendix-k} provides a structured Evaluation Card declaring \parbench{}'s measurement scope, assumptions, valid and invalid claims, and a recommended reporting protocol for future users. 

% \parbench{} is evaluation infrastructure rather than a new translation method. Its contribution is to make the measurement problem sharper: kernel-centric isolation avoids repository-level build confounds, declarative specs make correctness reproducible, augmentation probes robustness, and the harness exposes where translations fail. The initial results show that current LLMs have real but uneven parallel translation capability, with build adaptation, direction asymmetry, and multi-file coordination as central barriers to reliable use. The benchmark code, all 96 curated JSON evaluation specs, and per-record result files for all three models are included in the supplementary material to enable independent verification and extension.
