% §2.1 Declarative Specifications and Kernel-Centric Translation
%   - First: what a spec IS — JSON contract encoding identity, file
%     partitioning, build/run/verify config. Refer to appendix for schema.
%   - Then: the KEY DESIGN CHOICE — kernel-centric translation. LLM gets
%     only translation_targets; build infra stays fixed. This is what
%     separates \parbench{} from ParEval-Repo (0% vs 37–76%).
%   - Then: prompt construction — structured prompt with source code,
%     target API, build command, infrastructure context. Anonymization.
%   - Finally: cross-API argument/verification handling — union patterns
%     for full-program; target spec for kernel-only (OpenCL .cl).
%
% §2.2 Augmentation Engine
%   - First: WHY — training data contamination concern (Rodinia is
%     everywhere). Need to test whether success = memorization.
%   - Then: WHAT — six AST-level transforms (list them). All backed by
%     libclang. Preserve parallel structure, change surface features.
%   - Then: HOW MUCH — five levels (L0–L4), increasing density.
%   - Finally: VALIDATION — 80/87 specs pass all levels; 87/87 at L1–L2.
%     7 failures only in omp_target at high intensity.
%
% §2.3 Harness: Build, Run, Verify
%   - First: three-stage pipeline with short-circuit on failure.
%   - Then: what each stage does (build → run → verify).
%   - Then: WHY conjunctive verification matters — example of Gaussian
%     elimination passing compile+exit but failing output check.
%   - Finally: five failure classifications → enables Section 5.2 analysis.




\section{\parbench{} Framework}
\label{sec:framework}


To better evaluate LLM-based translation of parallel programming APIs, \parbench{} defines a controlled, kernel-centric setting that separates parallel kernel translation from repository-level challenges such as build-system reconstruction. The framework realizes this setting through four components: (1)~declarative specifications, (2)~an AST-driven augmentation engine, (3)~an evaluation pipeline that orchestrates LLM translation and records diagnostic failure categories, and (4)~a build-run-verify harness. Together, these components enable direct measurement of parallel API translation capability under end-to-end executability constraints while supporting controlled robustness analysis. Figure~\ref{fig:architecture} shows the overall architecture.

% \parbench{} evaluates LLM-based parallel code translation through automated, end-to-end correctness assessment. The framework has four components: (1)~declarative specifications that define correct execution for each kernel, (2)~a build-run-verify harness, (3)~an AST-driven augmentation engine for robustness testing, and (4)~an evaluation pipeline that orchestrates LLM translation and records failure categories. Figure~\ref{fig:architecture} shows the overall architecture.



\subsection{Declarative Specifications and Kernel-Centric Translation}
\label{sec:spec-schema}

Each benchmark instance in \parbench{} is encoded as a JSON specification (spec) that serves as a declarative contract defining what for ``correct execution'' means for that kernel--API variant. The spec encodes all information required to build, run, and verify a benchmark without manual intervention, enabling fully automated evaluation. The schema is formally defined in JSON Schema and enforced by an offline validator. A spec declares the kernel's identity, specifies the target API, and includes provenance details to ensure reproducibility by including the pinned repository and commit. The spec partitions source files into \texttt{prompt\_payload} (files provided to the LLM during translation), \texttt{support\_files} (build infrastructure such as Makefiles and shared headers, provided as read-only reference), and \texttt{verification\_only} (reference implementations never provided to the LLM).This field is central to the kernel-centric translation methodology described in Section~\ref{sec:eval-pipeline}. An example spec and the full structure are provided in Appendix~\ref{sec:appendix-a-schema}. 
\textbf{Build, run, and verification.} The \texttt{build} block specifies compiler commands, environment variables, and expected executable path. The \texttt{run} block defines input configurations with arguments and timeouts. The \texttt{verification} block declares an ordered list of verification strategies evaluated in conjunction. All must pass for an overall PASS verdict (Section~\ref{sec:harness}). Five strategy types are implemented: \texttt{exit\_code}, \texttt{stdout\_pattern}, \texttt{stdout\_exclude\_pattern}, \texttt{numeric\_comparison} (scalar tolerance check against a reference), and \texttt{file\_hash} (SHA-256 digest match).

\textbf{Baseline results.} The \texttt{baseline\_results} block records the expected output from running the original implementation on a reference platform, providing ground truth for correctness evaluation.

We have defined 96 such specs spanning five benchmark suites: 60 from Rodinia~\cite{Rodinia2009}, 4 from XSBench~\cite{XSBench2014}, 4 from RSBench, 3 from mixbench, and 25 from a curated subset of HeCBench~\cite{HeCBench2023}. Of these, 87 pass the baseline build-run-verify pipeline; the remaining 9 (\knownfail{} specs) fail due to platform-specific issues, such as CUDA~12 API deprecations, missing system libraries, or pre-existing runtime errors, and are excluded from evaluation (Section~\ref{sec:benchmark-curation}). Listing~\ref{lst:bfs_spec_appendix} in Appendix~\ref{sec:appendix-a} illustrates the spec structure using Rodinia's BFS kernel.

\subsection{Evaluation Pipeline: Kernel-Centric Translation}
\label{sec:eval-pipeline}

The key design choice is \textbf{kernel-centric translation}: the LLM is asked to modify only the \texttt{translation\_targets} files and produce functionally equivalent target-API code, while Makefiles, host scaffolding, and verification logic remain fixed. This isolates parallel API translation from repository reconstruction---the confound that drives ParEval-Repo's 0\% pass rate on applications above 133~SLoC~\cite{ParEvalRepo2025}. The evaluation pipeline constructs a structured prompt containing the source kernel code, the target API and output filenames, the build command, and read-only target infrastructure context, with prompt anonymization to reduce benchmark identity leakage. For cross-API translations, the pipeline handles argument and verification asymmetry automatically: full-program translations use a union of source and target verification patterns, whereas kernel-only translations, such as OpenCL \texttt{.cl} files, use the target spec directly. This design isolates evaluation of whether an LLM can correctly map parallel programming patterns from one API to another, independent of its ability to replicate project file organization. The full prompt template is in Appendix~\ref{sec:appendix-g}.

Each LLM response is parsed to recover the translated source files using a multi-tier extraction strategy (Appendix~\ref{sec:appendix-a}) that progressively relaxes matching criteria from explicit filename annotations to fuzzy and elimination-based heuristics. If any expected target file cannot be recovered or is empty, the task is classified as \extractionfail{}. To enable stratified analysis of translation difficulty, each source--target pair is classified by cardinality of the translation: \texttt{single\_file} ($1 source {\to}1 target$), \texttt{multi\_to\_single} ($N{\to}1$), \texttt{single\_to\_multi} ($1{\to}N$), or \texttt{multi\_to\_multi} ($N{\to}M$).

% FIX-3: Anonymization protocol description
%\textbf{Prompt anonymization.} To reduce the risk that benchmark-specific identifiers in the prompt trigger memorized translations, the evaluation pipeline applies six anonymization measures: (1)~the kernel name and description are omitted from the prompt entirely, (2)~source code comments are stripped from all files, (3)~source filenames are replaced with generic identifiers (``Source File 1'', etc.), (4)~target output filenames use a generic scheme (\texttt{translated\_0.ext}, \texttt{translated\_1.ext}, etc.), (5)~kernel names in build commands are replaced with generic placeholders, and (6)~support header and auxiliary files are renamed to generic identifiers. This anonymization is orthogonal to the AST-driven augmentation engine (Section~\ref{sec:augmentation-engine}), which modifies the source code \emph{structure}; prompt anonymization targets the surrounding \emph{metadata}. Full details appear in Section~\ref{sec:anonymization}.

% Commenting out as it looks confusing to talk about kernel-only translation and then removing false positives. It is
% understanble by now that we translate kernel + host code.
% \textbf{Kernel-only vs.\ full-program translation.} Not all cross-API translations require the LLM to rewrite the entire program. \parbench{} distinguishes \emph{kernel-only} translations --- where only the device kernel files are translated while the host code remains untouched --- from \emph{full-program} translations, where the LLM rewrites all source files, including host-side API calls. The distinction arises naturally from API architecture: translations targeting OpenCL produce only \texttt{.cl} kernel files, because the OpenCL programming model separates kernel source (compiled at runtime via \texttt{clBuildProgram()}) from host orchestration code. In contrast, CUDA-to-OpenMP translations require rewriting the entire source file, replacing device kernels, memory management, and launch syntax with OpenMP pragmas and CPU-side equivalents. \parbench{} detects kernel-only translations automatically by checking whether all \texttt{translation\_targets} in the target spec end with \texttt{.cl}. For kernel-only translations, the target spec's native run arguments and verification patterns are used directly (since the host binary is unchanged), whereas full-program translations use the cross-API argument and verification handling described above.

% \textbf{False-positive detection for kernel-only translations.} Kernel-only translations introduce a subtle verification challenge: the host program may swallow kernel compilation errors and still produce expected output. To guard against such false positives, the verification pipeline includes a post-verification scan of \texttt{stdout} for OpenCL kernel build failure signatures. Translations that pass exit-code and stdout-pattern checks but contain evidence of kernel compilation failure are downgraded from PASS to \verifyfail{}. This specificity filter reduces false positives with zero additional false negatives, since legitimate kernel executions do not produce \texttt{clBuildProgram()} error messages.

%\textbf{Optional self-repair.} The framework supports an iterative self-repair path (dashed arrow in Figure~\ref{fig:architecture}): on failure, the pipeline can feed failure-specific diagnostics back to the LLM for a corrected translation attempt. This capability is not exercised in the evaluation reported here; each sample is a single-attempt translation with no iterative repair (Section~\ref{sec:experimental-setup}).

\subsection{Augmentation Engine: Probing Robustness via Code Perturbations}
\label{sec:augmentation-engine}

Widely used suites such as Rodinia~\cite{Rodinia2009} are likely present in LLM training data, so success on unmodified source may reflect memorization. The augmentation engine generates semantically equivalent but syntactically novel variants by applying six AST-level transforms backed by the \texttt{libclang} API: condition operand swapping (e.g., \texttt{a < b} $\to$ \texttt{b > a}), arithmetic form conversion (e.g., \texttt{x += 1} $\leftrightarrow$ \texttt{x = x + 1}), scope-aware variable renaming, typedef expansion, pointer-to-array notation conversion e.g., \texttt{*(arr + i)} $\to$ \texttt{arr[i]}), and static function renaming. All transforms preserve parallel structure (thread indexing, synchronization, memory management) while changing surface-level code features most likely to be memorized verbatim. Parallelism-aware transforms (e.g., loop reordering) are excluded because they risk altering observable output.

Five augmentation levels (L0--L4) control density of applied transforms. L0 is unmodified source. L1 applies one transform to one site. L2--L4 apply increasing fractions of transforms and candidate sites ($f{=}0.33$, $0.66$, $1.0$). Baseline validation confirms semantic preservation: 80 of 87 non-\knownfail{} specs pass at all levels L1--L4, with zero correctness regressions. All 87 pass at L1--L2. The 7 high-intensity failures occur exclusively in \texttt{omp\_target} variants where aggressive variable renaming interferes with OpenMP target data-mapping clauses. This is a known limitation of the current transform guards. Level definitions and per-transform details are in Appendix Tables~\ref{tab:augmentation-levels} and~\ref{sec:appendix-e1}.

\subsection{Harness: Build, Run, Verify}
\label{sec:harness}

The harness is a three-stage pipeline where failure at any stage short-circuits subsequent stages. The \textbf{build stage} resolves the spec's working directory, substitutes platform-specific paths, and invokes compilation. The \textbf{run stage} executes the compiled binary with spec-defined arguments and timeout. The \textbf{verify stage} applies an ordered list of verification strategies in conjunction---all must pass for a \pass{} verdict. Five strategy types are implemented: \texttt{exit\_code}, \texttt{stdout\_pattern}, \texttt{stdout\_exclude\_pattern}, \texttt{numeric\_comparison}, and \texttt{file\_hash}(Section~\ref{sec:eval-pipeline}).

Conjunctive verification is essential. For example, a CUDA translation of Gaussian elimination compiles, runs to completion with exit code zero, and produces partial stdout---but omits the expected timing summary. This indicates a systematic translation defect rather than a transient failure. Compile-only~\cite{CodeRosetta2024} or exit-code-only verification would misclassify this as successful. Each outcome receives one of five failure classifications---\buildfail{}, \runfail{}, \verifyfail{}, \extractionfail{}, or \pass{}---enabling systematic failure-mode analysis (Section~\ref{sec:failure-taxonomy})(Section~\ref{sec:eval-pipeline}).


For kernels that produce quantitative output, stronger verification is available: \texttt{numeric\_comparison} extracts a scalar from stdout via a configurable regex and compares it against a reference value within a per-spec tolerance, while \texttt{file\_hash} verifies that a designated output file matches a reference SHA-256 digest. Of the 87 non-\knownfail{} specs, 7 use these stronger oracle; the remaining 80 rely on exit-code and stdout-pattern checks that capture each benchmark's own self-checking output (e.g., printed \texttt{PASS}/\texttt{FAIL} verdicts or expected timing summaries). Because the dominant failure mode is build-stage API adaptation (Section~\ref{sec:results}), the current oracle distribution is sufficient to classify the majority of outcomes; upgrading weak oracles to numeric verification for remaining specs is identified as future work (Section~\ref{sec:discussion}).

% A \texttt{--resume} flag allows batch evaluations to skip previously completed tasks, supporting append-only result accumulation without re-executing passing benchmarks.